Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy

### 2023-10-26 manuscript/manuscript.qmd
targets-ify manuscript, website, and readme
Suparna mega edits/updates

---
title: "Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy"
short-title: "Why Donors Donate"
date: "2023-09-09"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
  orcid: 0000-0003-3506-7590
- name: Marc Dotson
  affiliation: Brigham Young University
  url: https://marriottschool.byu.edu/directory/details?id=50683
  email: marc_dotson@byu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
abstract: |
  As space for civil society has closed around the world, transnational NGOs have faced a crisis of funding. We explore how NGOs have shifted from traditionally Northern funding sources toward grassroots private philanthropic money. How do individual donors in the US feel about donating to legally besieged NGOs abroad? Do legal restrictions on NGOs influence donors’ decision to donate? We use a conjoint survey experiment to argue that domestic political environments of NGO host countries influence preferences of private donors and that legal crackdowns on NGOs serve as a heuristic of organizational deservingness.
keywords:
  - philanthropy
  - civil society
  - conjoint experiments
  - donor heuristics
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/silent-skywalk>"
bibliography: references.bib
link-citations: true
---

@@

Over the last three decades, states have increasingly sought to crack down on civil society organizations. In Human Rights Watch's 2016 Annual Report, Executive Director Kenneth Roth argued that civil society was under more aggressive attack than at any time in recent memory. Policymakers have referred to this phenomenon as the "closing of civic space" and data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in nearly 60% of countries—111 in total [@CIVICUS:2017]. These attacks—most of which have occurred through restrictive legislation targeted against nongovernmental organizations (NGOs)—have adversely affected the ability of both domestic and international NGOs to operate and raise funds. Official aid funds have decreased substantially in countries that have imposed restrictive anti-NGO legislation [@ChaudhryHeiss:2018; @DupuyPrakash:2018] and official aid donors channel fewer funds to contentious issues and instead choose to fund "tame" causes [@ChaudhryHeiss:2018; @Bush:2015].

In response to this dramatic shift in the NGO funding landscape, international NGOs (INGOs) have increasingly turned to foreign individual donors to offset the loss of funding from government and foundation sources [@BanksHulmeEdwards:2015]. The amount of this private foreign aid is not insignificant—from 2000–2010, private giving by U.S. entities towards international causes doubled in real terms [@desaikharas, 505]. In 2016, U.S. private philanthropy towards international NGOs amounted to \$22.06 billion [@givingusa:2017].

This turn to private donors opens INGOs up to new dynamics in seeking out aid. Many organizations that had previously relied on governments and foundations had optimized their fundraising strategies for those kinds of organizations, but now need to tailor their fundraising attempts to private individuals, who behave differently from governments and foundations. Individuals have different preferences and rely on signals of effectiveness that differ from the benchmarks that large donor agencies rely on, including ratios of administrative overhead spending, financial transparency, accountability, and efficiency. These signals are are organizational characteristics that INGOs have direct control over. However, the political environment in the countries INGOs work in—including the global crackdown on civil society that is causing many organizations to seek out private funding in the first place—is a structural factor that INGOs can rarely influence on their own. How do organizational elements of private donor decision-making interact with broader structural limits on INGO fundraising?

Given the growing crackdown on INGOs across the world, how do individual donors in the U.S. feel about donating to legally besieged NGOs abroad? How do organizational-level factors such as financial transparency and accountability; funding sources, including private donations and government grants; and various different issue areas compare with more structural-level factors such as host country politics when deciding to donate?

While recent research on private foreign aid has examined funding decisions of donor NGOs themselves, we know less about the funding choices of the individuals responsible for much of these private aid flows [@desaikharas]. We argue that NGO-level organizational factors and domestic political environments of NGO host countries can influence preferences of foreign private donors by serving as heuristics to potential donors. Using a conjoint survey experiment, we find that private donors favor NGOs that voluntarily engage in transparency and accountability, prefer organizations that receive funding from many private individuals, rather than government or foundation grants, and that work on issues related to humanitarian and disaster relief rather than human rights and environmental advocacy. Importantly, we also find that structural factors play an important role in the private decision to donate—donors are significantly less likely to donate to INGOs that face government criticism or crackdown. Our results have important implications for international NGOs facing increased antagonism by government abroad and point to possible methods of adapting to these new regulations alienating their newfound donor base.

# Private foreign aid

Substantial research on nonprofit studies has looked at the motivation of individual donors, including altruism, reputational benefits and alignments with personal values [@BilodeauSlivinski:1997; @Wiepking:2010; @BekkersWiepking:2011]. However, this research has overwhelmingly looked at individual donor behavior and motivation for giving to local—rather than international—causes. On the other hand, the systematic study of private foreign aid is still in its infancy—there is an absence of expansive and high quality data [@greenhill2013], and most research has been restricted to elite, high net-worth donors [@USTrust2014].

We argue that understanding funding decisions of private citizens is important for two main reasons. Examining the organizational-level factors in a group that make the average donor increase their charitable-giving towards these groups can help NGOs better target their programming towards these donors. It can also help organizations market their activities more saliently. More broadly, understanding motivations of private citizens can help organizations and governments better assess claims about the benefits of privatizing foreign aid. Previous research has argued that private foreign aid does not have to deal with the same strategic and political considerations as official aid and therefore, may be in a better position to respond to recipients' needs on the grounds [@easterlywilliamson].

While recent research on private foreign aid has examined funding decisions of donor NGOs and foundations[^1], we know less about the funding choices of the individuals responsible for much of these private aid flows [@desaikharas]. While we do investigate the impact of individual-level factors on such private philanthropy targeted towards international causes in our larger project, in this paper we seek to understand the impact of organizational features of nongovernmental organizations and the structural factors in target countries in which they operate on donors’ decisions to engage in philanthropy.

@@

## Organizational heuristics: The impact of NGO organizational decisions on individual philanthropy

Existing research has found that donors respond to a host of heuristics and signals when deciding whether to donate to an NGO, since seeking out complete information about an organization's deservingness and efficiency is costly and time-consuming. One common signal for donors is organizational efficacy or the belief that an individual's donation will make a real difference. This can manifest itself through multiple mechanisms such as the amount an organization spends on overhead costs, whether an organization makes its finances public, or whether an organization participates in voluntary accountability practices, such as listing organizational details on websites like GuideStar and Charity Navigator.

Financial transparency, defined as the degree of completeness of information, provided by organizations to the public concerning its business activities is one such important organizational practice that can enhance donor trust and confidence in organizations [@Vacarro:2006; @vaccaromadsen]. Previous research has found that media exposés about charity mismanagement can generate negative reputational spillovers for the charitable sector [@gibelmangelman]. Therefore, being financially transparent and engaging in activities that further transparency can reduce actual or perceived information asymmetries between donors and charities, and increase donor perception of organizational efficacy.

@@

In addition to financial transparency, researchers and organizations, especially those auditing accountability practices, argue that accountability is another factor that can impact donor willingness to donate. While transparency is an organization's willingness to make critical data about the organization public, accountability is the willingness of the organization to explain its action to shareholders.

Explaining actions to stakeholders can be important because many NGOs, even when well-intentioned, can devote excessive resources towards marketing and overheads [@bowman:2006]. Private donors, subsequently, may be more likely to dislike NGOs that spend a large share of their revenues for unproductive purposes [@rose1982], or rather, a large share of resources on administration, management and fundraising. As @Tremblay-BoirePrakash:2017 point out, while individuals would like to have some assurance that their resources will be used appropriately and organizations are not spending too much on overhead, donors do not necessarily have the time to research charities thoroughly. Based on a survey experiment, they find that membership in a voluntary program ultimately does not influence people's willingness to donate. Thus, while seeking out explanations about NGO actions can be important to individual donors, they do not necessarily seek assurance through third-party programs/audits and charity watchdogs, but rather through word of mouth, personal scrutiny and local networks.

@@

Another heuristic for deservingness is the *issue area* an NGO engages in. Donors can have very different preferences regarding the nature of NGO activities that they would like to support [@AndreoniPayne:2003]. Organizations that address humanitarian issues such as relief and development have broader appeal to donors who might shy away from NGOs dealing with more political causes like human rights. Organizations focusing on the former may also be able to frame their activities as in need of greater support.

Perception of need, whether from personal experiences or acquired through media framing, can strongly influence the decisions of individuals to donate to relief efforts [@BekkersWiepking:2010, 6-8; @wiepkingleeuwen, 227-29]. This is especially the case when donors are faced with beneficiaries that fit the commonly accepted stereotypes of "victims" [@breezedean, 134]. Indeed, previous research has found that emergency funding, at least in the form of bilateral aid NGO, is more likely to be directed towards the senders of asylum-seekers and refugees [@czaikamayer; @bermeoleblang]. Therefore, donors may be more willing to donate to NGOs working on emergency response or relief focusing on victimized groups such as refugees or women and children in war.

On the other hand, donors may be less willing to support NGOs focusing on contentious issue areas or programs that are not compatible with host government preferences. While NGOs working on issues like health, education and humanitarian relief rarely challenge the government [@Bush:2015; @Kiai:2017], those working on more contentious areas such as human rights, environmental advocacy, or election monitoring can pose political and economic challenges to regimes [@Chaudhry:2016]. These organizations are often viewed as politically motivated outsiders and run a higher risk of getting expelled [@DPR2015], increasing the possibility of wasting donor resources. Further, unlike the former causes, their programmatic output is harder to measure and quantify, which is another factor that is important to donors [@Bush:2015].

>**H~2~**: If NGOs work in less contentious issue areas (emergency response and refugee relief), then individual private donors will be more likely to donate to them over more contentious issue areas (environment and human rights).

The *source of NGO funding* can also influence donor behavior. Government grants to NGOs can often crowd out private funding [@AndreoniPayne:2003]. Donors are less likely to contribute to NGOs that rely heavily on government grants since their donation might feel more impersonal and less needed, while donors are more likely to give to organizations that rely primarily on private donations given that they may feel personally invested in the organization and perceive their contribution mattering. Government grants may also imply lack of independence from government which can reduce the efficiency of an organization [@Stroup:2012].

>**H~3~**: If NGOs are funded primarily by numerous small private donors, then individual private donors will be more likely to donate to them compared to NGOs that are funded by a handful of wealthy private donors and government grants.

Each of these signals is *organizational*. NGOs typically have direct control over their organizational practices, mission and issue area, and funding sources. We theorize that organizational characteristics like managerial practices, issue area, and funding sources all influence the decision to donate to an NGO (see @fig:big-dag). At the same time, however, we propose that structural characteristics like an NGO's host-country political and legal environment can also serve as donor heuristics.

## Structural heuristics: The impact of anti-NGO crackdowns on individual philanthropy

NGOs have less control over *structural* issues, and the role of political and legal institutions in donor decision making has been understudied. However, over the past decade international NGOs have become increasingly politicized and legally restricted around the world. NGO registration is not inherently restrictive, as governments frequently use regulations to streamline NGO-state relationships. Restrictive NGO laws, on the other hand, are designed to limit organizational programming by imposing barriers to entry, funding, and advocacy. These anti-NGO laws may signal to donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and crack down on them. Therefore, we expect that government restrictions abroad will increase respondents' likelihood of donating to the organization.

>**H~4~**: If NGOs face legal crackdowns abroad, then individual donors will be more likely to donate to them.

@@

Identifying the causal link between structural factors and donor behavior is more difficult than measuring the link between organizational factors and donor behavior, however, because (1) NGOs have less direct control over their political environments, and (2) organizational characteristics like issue area and funding can sour the NGO-government relationship (as we explain in further detail below, governments are more likely to crack down on NGOs that address human rights and that receive government funding). @fig:big-dag helps disentangle this complex relationship. Organizational factors (**O**, or organizational practices like financial disclosure and accountability practices; **I**, or issue area; and **F**, or funding) all influence the decision to donate (**D**). Structural factors like an NGO's relationship with its host government (**G**) also influence donation behavior, but the relationship itself is shaped by both funding and issue area. As such, legal crackdowns do not occur in isolation. Countries are more likely to repress NGOs when organizational missions are in tension with government preferences and when organizations receive substantial funding from abroad [@DPR2015]. Therefore, the same organizational characteristic that can increase the likelihood of crackdown can also simultaneously serve as a donor heuristic and influence perceptions of NGO deservingness.

Donors therefore may show increased willingness to donate to NGOs facing government crackdown/criticism working in less contentious issue areas (emergency response and refugee relief) over more contentious issue areas (environment and human rights). Those working on contentious issues may be expelled or shut down, which would waste donor resources and make it less likely that they donate to these organizations. Similarly, the combination of legal crackdown and funding may reshape donor perceptions of deservingness. Private donors may be more likely to punish government-funded NGOs that face restrictions. This is because legal trouble with the group's main funder could fuel private donor mistrust.

Finally, the combination of financial transparency and government crackdown may increase donor willingness to support besieged NGOs. While concerns about NGO mismanagement and lack of financial transparency may signal concerns about an NGO's operations and its ability to abide by the host country's regulatory environment, the presence of such transparency may instead convince donors that government targeting of NGOs may be ill-intentioned rather than simply a manifestation of financial oversight. Donors may view NGO legislation as a sign of governments looking to crack down on groups that seek to keep governments accountable, rather than merely auditing and punishing groups that violate routine regulations.

> **H~5a~**: Donors will show increased willingness to donate to NGOs facing government crackdown/criticism working in less contentious issue areas (emergency response and refugee relief) over more contentious issue areas (environment and human rights).
>
> **H~5b~**: Donors will show increased willingness to donate to NGOs that are facing government crackdown and are funded primarily by numerous small private donors.
>
>**H~5c~**: Donors will show increased willingness to donate to NGOs that are facing government crackdown and are funded primarily by numerous small private donors and work in less contentious areas (emergency response and refugee relief).
>
>**H~5d~**: Donors will show increased willingness to donate to NGOs that are criticized by the government/under government crackdown when they are also financially transparent and work in less contentious areas (emergency response and refugee relief) and are funded by numerous small donors.

@@

After collecting baseline information on respondent demographics and attitudes toward charity and voluntarism, we presented respondents with four randomly selected international nonprofit organizations with four randomly assigned features.^[See the appendix for a copy of the survey.] We included organizations that are associated with a range of contentious and noncontentious issues: Amnesty International, Greenpeace, Oxfam, and the International Committee of the Red Cross. We then displayed various combinations of features, including one of four issue areas (emergency response, environmental advocacy, human rights advocacy, and refugee relief), one of two organizational practices (financial transparency and third-party accountability audits), one of three funding sources (many small private donations, a handful of wealthy private donors, and government grants), and one of three relationships with host governments (friendly, criticized by the government, and under government crackdown). We presented twelve sets of four hypothetical combinations of organizational features, and because our partial fractional factorial design resulted in 288 (4 × 4 × 2 × 3 × 3) possible combinations of organizations and features, no single respondent was offered every combination.

@@

# Discussion and conclusion

In general we find support for our hypotheses. With the turn towards private philanthropy in the face of the global crackdown on civil society, INGOs need to be aware of the dynamics of individual donor preferences and how those preferences interact with their managerial and organizational decision. Additionally, structural factors like political trends in INGO host countries influence donor decision-making. In aggregate, donors are less likely to give to organizations that face government criticism and crackdown, which could be a worrying sign for INGOs working on contentious issues that are more likely to elicit negative government responses. However, for the subgroup of respondents with personal involvement in advocacy, government crackdown serves as a positive heuristic instead, likely because those involved in advocacy understand how the relationship between governments and charities can be fraught.

>>>>

---
title: "Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy"
short-title: "Why Donors Donate"
date: "2023-09-09"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
  orcid: 0000-0003-3506-7590
- name: Marc Dotson
  affiliation: Brigham Young University
  url: https://marriottschool.byu.edu/directory/details?id=50683
  email: marc_dotson@byu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
abstract: |
  Prior research on nonprofit fundraising has primarily examined how financial transparency and organizational accountability influence individual donor behavior. Over the past two decades, however, an increasing number of host governments have cracked down on non-governmental organizations (NGOs). QUESTION. We test these questions by administering a series of conjoint experiments in the US that examine...We find.... This suggests...
keywords:
  - philanthropy
  - civil society
  - conjoint experiments
  - donor heuristics
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/silent-skywalk>"
bibliography: references.bib
link-citations: true
---

@@

```{r libraries-data}
library(tidyverse)
library(targets)

# Targets stuff
tar_config_set(store = here::here('_targets'),
               script = here::here('_targets.R'))

invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
```

Over the last three decades, governments globally have increasingly sought to crack down on NGOs (CITE). Repressive governments have repressed these organizations using both violence, as well as administrative crackdown, which create barriers to funding and advocacy for a variety of NGOs. (CITE) Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" (CITE) and data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in nearly 60% of countries—111 in total (UPDATE FROM NEW CIVICUS DATA). These attacks—most of which have been implemented through anti-NGO laws—have adversely affected the ability of both domestic and international NGOs to operate and raise funds. Official aid funds have decreased substantially in countries that have imposed restrictive anti-NGO legislation [@ChaudhryHeiss:2018; @DupuyPrakash:2018] and official aid donors channel fewer funds to contentious issues and instead choose to fund "tame" causes [@ChaudhryHeiss:2018; @Bush:2015, HERROLD, SPRINGMAN]. In response to this dramatic shift in the NGO funding landscape, international NGOs (INGOs) have increasingly turned to foreign individual donors to offset the loss of funding from government and foundation sources [@BanksHulmeEdwards:2015].

The amount of private foreign aid is not insignificant—from 2000–2010, private giving by U.S. entities towards international causes doubled in real terms [@desaikharas, 505]. In 2016, U.S. private philanthropy towards international NGOs amounted to \$22.06 billion [@givingusa:2017]. Most notably, a large portion of this growth was driven by individuals making small-scale donations. In 2017, individuals gave US\$286.7 billion or approximately 70% of total giving to international affairs. This further increased to US\$427.71 billion in 2018, 85.7% of which was through individuals (Giving USA, 2019). This turn to private donors opens INGOs up to new dynamics in seeking out aid. Many organizations that had previously relied on governments and foundations had optimized their fundraising strategies for those donors, but now need to tailor their fundraising strategies to private individuals, who behave differently from governments and foundations.

Individuals have different preferences and rely on signals of effectiveness that differ from the benchmarks that large donor agencies rely on, including financial transparency and accountability. While INGOs have control over these organizational-level factors that motivate donations, the political environment in the host countries INGOs work in—including the global crackdown on civil society that is causing many organizations to seek out private funding in the first place—is a structural factor that INGOs can rarely influence on their own. How do organizational elements of private donor decision-making interact with broader structural limits on INGO fundraising? How do individual donors in the U.S. feel about donating to legally besieged NGOs abroad? How do organizational-level factors such as financial transparency and accountability compare with more structural-level factors such as host country politics when deciding to donate?

Although a robust body of work on NGOs and donor behavior examines the motivation of public donors, private foundations and elite donors (CITE), we are yet to examine the impact of closing civic space, especially individual private donors, who are likely to be important funders in the face of increased government repression. Existing research shows that organizational accountability and financial transparency play an important role in motivating individual donors to donate to NGOs (CITE). In this article, we extend this literature by evaluating how these organization-level characteristics influence private donor behavior in an era of shrinking civic space. In considering this question, we engage with a growing body of research on the consequences of closing civic space on NGOs and donors (CITE). We argue that NGO-level organizational factors and domestic political environments of NGO host countries serve as heuristics and can influence foreign private donors' preferences. We fielded a conjoint experiment to a nationally representative sample of X adults in the US. A conjoint design allows us to directly compare the relative impact of host government repression of NGOs vis-a-vis an organization's financial transparency and accountability on individual donor preferences. The result of these experiments show both financial transparency and accountability increases individuals' willingness to donate NGOs. Learning about host government criticism and crackdown against NGOs decreases the likelihood of donation by itself; however, organizational features such as financial transparency and accountability protect against this dampening effect. INSERT LINE ABOUT EFFECT SIZE.

These findings make two main contributions. While recent research on private foreign aid has examined funding decisions of donor NGOs and foundations[^1], we know less about the funding choices of the individuals responsible for much of these private aid flows [@desaikharas]. The systematic study of private foreign aid is still in its infancy—there is an absence of expansive and high quality data [@greenhill2013], and most research has been restricted to elite, high net-worth donors [@USTrust2014]. Our research makes a significant contribution to research on individual giving and indicates the promise and limits of different framing appeals on individual donor preferences.

Second, research shows that NGO leades often have difficulty in attracting new donors and also find it challenging to moviate current donors (Cao 2015). Examining the organizational-level factors in a group that make the average donor increase their charitable-giving towards these groups, especially when facing restricted civic space abroad, can help NGOs better target their programming and frame their funding appeals towards these donors. Our results thus have important implications for international NGOs facing increased antagonism by government abroad and point to possible methods of adapting to worsening or shrinking civic environments.

Below, we summarize existing research on determinants of individual giving to foreign NGOs, laying out our expectations regarding how financial transparency, accountability shape individuals' philanthropic preferences in an environment of closing civic space. We then describe our conjoint experiment and and present the results. We conclude with implications for NGOs working in repressive countries.

@@

# What determines individual donor behvior?

Substantial research on NGOs has looked at the motivation of individual donors, including altruism, reputational benefits and alignments with personal values [@BilodeauSlivinski:1997; @Wiepking:2010; @BekkersWiepking:2011]. However, this research has overwhelmingly looked at individual donor behavior and motivation for giving to local—rather than international—causes (CITE -2). Understanding the latter is important as philanthropy to international causes can be harder to motivate because the number of recipients is larger and further removed from the donor (Casale & Baumann, 2015, 100). Further, as Micklewright and Schnepf (2009) point out, motivations for donating to ongoing international causes differ from donating money to natural disasters or emergency international causes. Under ordinary circumstances, “donors are more likely to donate to a charity operating locally than to a charity providing identical service abroad” (Tremblay-Boire & Prakash, 2017, p. 644). However, we lack adequate data and theory regarding philanthropy towards INGOs.

The third wave of democratization, the collapse of the Soviet Union, and the spread of a "pro-NGO norm" led to an increased role and visibility for NGOs. Globally, a large amount of resources were channeled to NGOs focused on service provision or development because many public and private donors perceived them as more efficient than states with weak institutions and poor governance; there was a high risk that such states would use aid transfers inefficiently (Dietrich 2013).

But as INGOs became increasingly powerful actors in global policy circles, they faced a number of criticisms. Many organizations struggle with the conflicting imperatives of being large organizations that can operate efficiently at scale, while maintaining the grassroots contacts and consensus building that contributed to their early successes.\footnote{\cite{mitchell2020between}, 2; \cite{jalali2013, bano2012, rahman2006}.} Many nonprofit scandals involve some form of financial wrongdoing (Prakash and Gugerty 2010). Globally, around 5\% of revenue is lost annually due to fraud (Lamothe et al 2022). Misappropriation of assets and embezzlement schemes are among the most common forms of financial fraud (Lamthoe). These criticisms and concerns, in turn, have led donors to demand greater NGO responsiveness, especially in terms of financial transparency and accountability.

Donors respond to a host of heuristics and signals when deciding whether to donate to an NGO, since seeking out complete information about an organization's deservingness and efficiency is costly and time-consuming (Croson & Shang, 2011; Tremblay-Boire & Prakash, 2017). Nonprofits fundraising internationally may therefore rely on framing to control perceptions. Below, we theorize the effect of different frames surrounding a country’s legal environment and nonprofit organizational characteristics on individual donor preferences.

## Organizational heuristics: The impact of NGO organizational practices on individual philanthropy

In theory, donors should scrutinize each aspect of an NGO’s organizational structure and programmatic performance prior to donation, but this rarely happens in practice (Tremblay-Boire & Prakash, 2017). Instead, individuals rely on heuristics to simplify their decision-making.

One common signal for donors is organizational efficacy or the belief that an individual's donation will make a real difference. This can manifest itself through multiple mechanisms such as the amount an organization spends on overhead costs, whether an organization makes its finances public, or whether an organization participates in voluntary accountability practices, such as listing organizational details on websites like GuideStar and Charity Navigator. An NGO's financial transparency, defined as the degree of completeness of information, provided by organizations to the public concerning its business activities, is one tangible heuristics that donors use to make their decisions. NGOs can engage in financial transparency in a variety of ways such as distribution of audiented financial statements, third party intermediaries. Reserach shows that transparency is associated with greater donations, lower debt and better governance (Harris and Neely 2021, Saxton et al 2012). It can also enhance donor trust and confidence in organizations [@Vacarro:2006; @vaccaromadsen].

Previous research has found that media exposés about charity mismanagement can generate negative reputational spillovers for the charitable sector [@gibelmangelman]. Diversions, or the unauthorized use of an organization's assets, including embezzlement and theft can result in a decrease in donatons. This effect becomes even stronger with media coverage of diversion (Harris et al 2023). Organizational financial transparent can reduce actual or perceived information asymmetries between donors and charities, this potentially increasing donor likelihood of donating to an INGO.

@@

Accountability is another factor that can impact donor willingness to donate. Accountability is necessary to promote public trust in NGOs. While transparency is an organization's willingness to make critical data about the organization public, accountability is the willingness of the organization to explain its action to shareholders. It also encompasses how well an organization's mission is aligned with its reources and how it manages resources to optimize performance (Sloan 2009).

Explaining actions to stakeholders can be important because many NGOs, even when well-intentioned ones, can devote excessive resources towards marketing and overheads [@bowman:2006]. Private donors, subsequently, may be more likely to dislike NGOs that spend a large share of their revenues for unproductive purposes [@rose1982], or rather, a large share of resources on administration, management and fundraising. As @Tremblay-BoirePrakash:2017 point out, while individuals would like to have some assurance that their resources will be used appropriately and organizations are not spending too much on overhead, donors do not necessarily have the time to research charities thoroughly. Lack of accountability is associated with lesser public trust, reputation, perceived quality and donation behavior (Becker 2018).

To judge organizational accountabiltiy, donors may use ratings by watchdog organizations like Charity Navigator and GuideStar as heuristics to guide their decisions. Research shows that initiatives to address accountability gaps by these organizations can affect donations to organizations in a significant manner (Sloan 2009, Gordon, Knock and Neely).

@@

## Structural heuristics: Host country conditions

In addition to organizational factors discussed above, political contexts in NGO host countries, especially their environments for civil society organizations, can also influence donor preferences for donating to foreign NGOs. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations (Glasius et al, Schmidt et al, Chaudhry, Heiss, DPR). As a result, NGOs must spend more time, effort, and resources to ensure their survival -- at the expense of pursuing their missions. Some NGOs have adapted to their repressive environments by recreating their organizations and changing their mission to avoid directly confronting the government. Analyzing the impact of anti-NGO laws in Bangladesh and Zambia, scholars have similarly found that NGOs with a broad mission ended up restricting themselves to service work and giving up on their advocacy work. Those that continued to engage in transnational advocacy changed the targets of their advocacy, the issues they focused on, and the language they used. For instance, many NGOs stopped lobbying the central government and focused their advocacy efforts on lower levels of government.\footnote{\cite{fransen2021tempering}.} Similarly, in Russia, many NGOs working with foreign partners and lobbying the central government no longer consider it an effective strategy.\footnote{\cite{sundstrom2022evolution}.} Overall, fewer groups may end up focusing on contentious causes, instead focusing on regime-compatible causes.

The changes in civic environments impacting NGO activity have also influenced official donor responses. In response to the passage of these laws, scholars have found that while multilateral donors do not reduce aid following the passage of anti-NGO laws, they are associated with a 32\% decline in bilateral aid inflows.\footnote{\cite{DupuyPrakash:2018}.} Further, advocacy-oriented donors -- i.e., those that fund democracy and civil society promotion activities -- reduced their spending by 74\%, but did not cut their spending on development projects such as education, health, water, and sanitation.\footnote{\cite{right2022advocacy}. However, previous literature has established that individual private donors may not necessarily use the same metrics as official donors (CITE). Private foreign aid does not have to deal with the same strategic and political considerations as official aid and therefore, may be in a better position to respond to recipients' needs on the grounds [@easterlywilliamson]. These anti-NGO laws may signal to private donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and crack down on them. Therefore, we expect that government restrictions abroad will increase respondents' likelihood of donating to the organization.

>**H~2~**: If NGOs face legal crackdowns abroad, then individual donors will be more likely to donate to them.

@@

Identifying the causal link between structural factors and donor behavior is more difficult than measuring the link between organizational factors and donor behavior, however, because (1) NGOs have less direct control over their political environments, and (2) organizational characteristics like financial transparency and accountability can sour the NGO-government relationship -- governments may be more likely to target NGOs that lack financial transparency and accountability. @fig:big-dag helps disentangle this complex relationship. Organizational factors (**O**, or organizational practices like financial disclosure and accountability practices; **I**, or issue area; and **F**, or funding) all influence the decision to donate (**D**). Structural factors like an NGO's relationship with its host government (**G**) also influence donation behavior, but the relationship itself is shaped by both transparency and accountability. As such, legal crackdowns do not occur in isolation. Countries are more likely to repress NGOs when organizational features are in tension with government preferences and when organizations receive substantial funding from abroad [@DPR2015]. Therefore, the same organizational characteristic that can increase the likelihood of crackdown can also simultaneously serve as a donor heuristic and influence perceptions of NGO deservingness.

Donors therefore may show increased willingness to donate to NGOs facing government crackdown/criticism while being financial transparent and accountable. While concerns about NGO mismanagement and lack of financial transparency may signal concerns about an NGO's operations and its ability to abide by the host country's regulatory environment, the presence of such transparency and accountability may instead convince donors that government targeting of NGOs may be ill-intentioned rather than simply a manifestation of financial oversight. Donors may view anti-NGO laws and inhospitable civic environments as a sign of governments looking to crack down on groups that seek to keep governments accountable, rather than merely auditing and punishing groups that violate routine regulations.

Since we are running a conjoint experiment, the interactions between the main treatment variables described above is simply an exploratory analysis. Importantly, when testing these hypotheses, we do not assume that individual donors have perfect knowledge about an NGO or the country an NGO operates in. Moreover, the average donor will not know exactly what “crackdown” entails for any given organization (i.e., does it mean an NGO was expelled, or that its assets were frozen, or that it received a fine?). For the sake of this experiment, this ambiguity is not a central concern as the heuristics we explore here are explicitly simplified short- cuts to donor information-gathering. Knowing that an organization is financially transparent (even if donors are not familiar with specific kinds of requirements regarding transparency) or knowing that an NGO faces legal trouble abroad should be enough to trigger the effect of these frames on donor preferences. Moreover, although many NGOs make information about their operations accessible through annual reports, it is time-consuming to track down that information. In each of these hypotheses, we assume that donors will not spend additional effort to research the exact details concerning crackdown, transparency and accountability. Instead, donors rely on signals and heuristics about the organiza- tion to shape their preferences and behavior (Szper & Prakash, 2011).

@@

After collecting baseline information on respondent demographics and attitudes toward charity and voluntarism, we presented respondents with four randomly selected international nonprofit organizations with four randomly assigned features.^[See the appendix for a copy of the survey.] We included organizations that are associated with a range of contentious and noncontentious issues: Amnesty International, Greenpeace, Oxfam, and the International Committee of the Red Cross. We chose these 4 organizations as the represent the core sectors of activity according to Charity Navigator (Environment, Human Rights, Humanitarian Relief, and Development). We then displayed various combinations of features, including one of four issue areas (emergency response, environmental advocacy, human rights advocacy, and refugee relief), one of two organizational practices (financial transparency and third-party accountability audits), one of three funding sources (many small private donations, a handful of wealthy private donors, and government grants), and one of three relationships with host governments (friendly, criticized by the government, and under government crackdown). We presented twelve sets of four hypothetical combinations of organizational features, and because our partial fractional factorial design resulted in 288 (4 × 4 × 2 × 3 × 3) possible combinations of organizations and features, no single respondent was offered every combination.

@@

# Conclusion
Over the last two decades, governments globally have systematically controlled and repressed NGOs using both violent tactics and administrative means over the last three decades -- a phenomenon known as `closing' or `shrinking' civic space. This trend of closing civic space has important implications for local and international NGOs as well as donors. Recent research has analyzed the impact of this crackdown on public or official aid donors. However, we know relatively little about how state repression of NGOs affects the preferences of foreign private donors. The responses of elite donors and foundations to government crackdown on civil society organizations remains a ripe area for research. Unlike state donors, individual and elite donors may not necessarily withdraw or reduce support for NGOs facing harassment abroad. This is important because private philanthropy towards organizations working in international affairs continues to grow despite inflation -- from 2021 to 2022, individual giving to nonprofits in international affairs grew by over 10.9% in the US (https://philanthropy.iupui.edu/news-events/news-item/giving-usa:-total-u.s.-charitable-giving-declined-in-2022-to-$499.33-billion-following-two-years-of-record-generosity.html?id=422).

Given the increasing hurdles faced by these nonprofits, how do individual donors in the U.S. feel about donating to legally besieged NGOs abroad? How do organizational-level factors such as financial transparency and accountability compare with more structural-level factors such as host country politics when deciding to donate? Using a conjoint experiment with a nationally representative sample of the U.S. population, we find COMPLETE.

These results have important implications for NGO operations, fundraising and survivability at a time when a number of INGOs are dealing with hostile host governments. The crackdown is even more challenging for local NGOs in Global South countries, where much government and public support is channeled via service delivery organizations. Local groups tackling contentious issues may not be able to raise funds domestically, as publics may prefer donating to service organizations focusing on health, education, and poverty.\footnote{\cite{absar2017will}.} Even local philanthropists may be deterred from giving to organizations focusing on contentious causes such as advocacy, media freedom, and anti-corruption initiatives due to poor tax incentives or a fear of retribution.\footnote{\cite{Brechenmacher:2017}; \cite{DPR2015}; \cite{Baoumi:2016}.} Our results highlight the importance of organizational characteristics like transparency and accountability even in an era of closing civic space. By finding new ways to communicate about organizational transparency and accountability to potential donors, NGOs can help overcome individual donors' concerns about host governments' criticisms of and crackdown on these groups.

Future research should test how well these results map on to elite individual donors, not just the average private donor, as well as determine the extent to which these preferences translate into behavior.
